---
title: "Producing plots"
chapter: true
weight: 20
pre: "<b>3. </b>"
---
### Chapter 3

# Producing plots


## Plotting a "gapminder" graph

In order to avoid getting bogged down on the syntax of ggplot2, a function to produce a "gapminder" plot is provided in the `workshopFunctions.R` file, which we load at the start of our app (See [git:02_loaddata]() ).  This uses ggplot2 to produce a graph, deals with setting fixed axes, consistent colours etc.

Having loaded the data, libraries and functions, we can produce a gapminder plot in R using the following code:

```{r, include=FALSE}
# Load the required libraries
library(ggplot2)
library(dplyr)
library(shiny)

gapminder <- readRDS("../../gapminder.rds")
source("../../workshopFunctions.R")

```


```{r}
gapminder %>% 
  filter(year == 2000) %>% 
  produceGapminderPlot()

```


## Exercise:

Add another `filter` to the example to only show countries that are in Europe and Africa

## Solution:
{{%expand "Click here" %}}
```{r}

gapminder %>% 
  filter(year == 2000) %>% 
  filter(continent %in% c("Europe", "Africa")) %>% 
  produceGapminderPlot()
```


{{% /expand%}}

{{% notice tip %}}
Further examples of using the plotting functions are given in the `codeExamples.R` file, which is in `~/mawdsley`
{{% /notice %}}
